---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r global_options, include=FALSE, }
knitr::opts_chunk$set(options(scipen=999))
```

## Loading and preprocessing the data
```{r include=FALSE}
library(dplyr)
library(ggplot2)
```
Load the dataset. Create copy without rows that don't a step count (steps = NA):
```{r}
dataset <- read.csv(unz("activity.zip", "activity.csv"),na.strings = "NA")
dataset_noNA <- dataset[!is.na(dataset$steps),]
summary(dataset)
```


## What is mean total number of steps taken per day?
Calculate the total number of steps made each day and list them out in an easily readable way. Then make a histogram of the values:
```{r}
#Sum steps over each day
day_sums <- aggregate(x=list(steps=dataset_noNA$steps), by=list(date=dataset_noNA$date), FUN=sum)
#Print step sums
as.numeric(day_sums$steps)
#Plot values in histogram
hist(day_sums$steps, xlab = "Daily Steps (In 5000 Step Increments)", ylab="# Of Days", main="") 
```

Calculate the average number of steps taken per day:
```{r}
avg <- sum(day_sums$steps) / length(day_sums$steps)
med <- median(day_sums$steps)
```

The Average # of Steps Per Day is: **`r avg`** steps   
The Median # of Steps Per Day is: **`r med`** steps

## What is the average daily activity pattern?
Calculate the average number of steps made during each 5-min interval across all days and provide a summary of values. Then make a histogram of the values:
```{r}
#Average steps over each 5-min interval
int_avg <- aggregate(x=list(steps=dataset_noNA$steps), by=list(interval=dataset_noNA$interval), FUN=mean)
#Print summary of step averages
summary(int_avg$steps)
#Plot values in time series plot
plot(x=int_avg$interval, y=int_avg$steps, type="l", xlab = "Time of day / 5-min", ylab="Steps", main="") 
```

Find the interval where the highest number of steps were taken on average across all days:
```{r}
int_max <- which.max(int_avg$steps)
```

The 5-min interval with the maximum steps taken, averaged across all days, is interval \#**`r int_max`**

## Imputing missing values
Store the indicies of rows with NA step values in original raw dataset:
```{r}
NA_i <- which(is.na(dataset$steps))
```
There are **`r length(NA_i)`** rows with NA steps.

Create a copy of the raw dataset and replace the NA step values in the copy with the intervals average value 
```{r}
#Create copy of raw dataset
dataset_imp <- dataset
#Function to return up interval step average
MyReplace <- function(index){
  int_avg$steps[int_avg$interval==dataset$interval[index]]
}
#Replace all NA values with interval step average
for(item in NA_i){
  dataset_imp$steps[item] <- MyReplace(item)
}
```

Using new imputed dataset, sum steps across each day, show results in a histogram and calculate the mean and the median for both the imputed and raw dataset:
```{r}
#Calculate sum of steps taken each day
day_sums_imp <- aggregate(x=list(steps=dataset_imp$steps), by=list(date=dataset_imp$date), FUN=sum)
#Create a histogram of steps from the imputed data
hist(day_sums_imp$steps, xlab = "Daily Steps (In 5000 Step Increments)", ylab="# Of Days", main="")
#Calculate mean and median for raw and imputed datasets
summary_raw_data <- summarise(day_sums, Median=median(steps), Mean=mean(steps))
summary_imp_data <- summarise(day_sums_imp, Median=median(steps), Mean=mean(steps))
```
The change in **mean** between the raw data and the imputed data is: **`r summary_imp_data$Mean-summary_raw_data$Mean`**  
The change in **median** between the raw data and the imputed data is: **`r summary_imp_data$Median-summary_raw_data$Median`**

Because NA values appeared across whole days, imputing average values for the NAs did not change the mean value of total daily steps, and changed the median values minimally.

## Are there differences in activity patterns between weekdays and weekends?
Create new column identifying if day fell during the week or on a weekend:
```{r}
#Function which takes a list of days of the week and creates a new list labelling weekdays vs. weekends 
MyWeekday <- function(x){
  #Creates new list of TRUE and FALSE (True = weekend, False = weekday)
  tmp <- x %in% c("Saturday","Sunday")
  #Replace true and false with weekend and weekday
  tmp[tmp==TRUE] <- "weekend"; tmp[tmp==FALSE] <- "weekday"
  #Return new list
  tmp
}
#Add list created with above function to the imputed dataset
dataset_imp <- mutate(dataset_imp, Weekday = MyWeekday(weekdays(as.Date(dataset_imp$date))))
#Check to see the levels of the new column
table(dataset_imp$Weekday)
```

Calculate the average steps across invervals and day type (weekend vs weekday) and plot:
```{r}
#Makes a long list of step averages for each interval/weekday combination (288 x 2 = 576 rows)
averages <- aggregate(steps ~ interval + Weekday, data = dataset_imp, FUN=mean)
#Plots two time series plots of the average steps taken for each interval. One plot for weekdays one plot for weekends
ggplot(averages, aes(x = interval, y = steps)) + geom_line() + facet_grid(Weekday~.) + xlab("Interval") + ylab("Avg. Steps")
```
